---
title: "Comparing Period TFR for 2 groups"
author: "Carl Schmertmann"
date: 10 Nov 2019
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen=999) # no scientific notation
library(tidyverse)

detail_year = 1995  # year for detailed US comparisons
```

## TFR for one group

Suppose that there are $A$ possible single-year reproductive ages $x=1\ldots A$ in which individuals may have children. At some of these ages fertility rates may be zero for some groups (for example, for women at $x > 50$).

Suppose also that there are a total of $B$ births in a given year to some group of interest (such as "all women", "all men", "Asian-American men", "residents of Montana" ...).  Call $\pi_x$ the fraction of the $B$ births that happened to group members who were age $x$. That is, births disaggregated by age are 
$$
B_x = B \:\pi_x \quad x=1,2,\ldots,A
$$ 
If there were $N_x$ group members at age $x$ at mid-year, then period $TFR$ for this group is

\begin{equation}
\label{eq:mainTFR}
TFR = \sum_x \frac{B_x}{N_x} = B \: \sum_x \pi_x \left(\frac{1}{N_x}\right) = B \: E_{\pi} \left( N^{-1}_x\right) 
\end{equation}

where the expectation in (\ref{eq:mainTFR}) is taken over the observed distribution of parents' ages for this year's births. 


## Comparing TFR for 2 groups

If there are two groups then the ratio of their period $TFR$s is

\begin{equation}
\label{eq:TFRratio}
\frac{TFR_1}{TFR_2} = 
\frac{B_1 \: \cdot \: E_{\pi_1} \left(  N^{-1}_{1x}\right)} 
     {B_2 \: \cdot \: E_{\pi_2} \left(  N^{-1}_{2x} \right)}
\end{equation}

In some cases (for example when the groups are "women" and "men") the total births are the same for both groups. For those cases:
\begin{equation}
\label{eq:TFRratiosameB}
\frac{TFR_1}{TFR_2} = \frac{E_{\pi_1} \left( N^{-1}_{1x}\right)} 
                           {E_{\pi_2} \left(  N^{-1}_{2x} \right)}
\end{equation}

Using the results derived in the Appendix below, the ratio in (\ref{eq:TFRratiosameB}) can be approximated fairly well by

$$
\frac{TFR_1}{TFR_2} \approx  
               \frac{N_{2,{m_2}}} 
                    {N_{1,{m_1}}}  
$$
where $m_1$ and $m_2$ are the mean ages of parents in the two groups for this year's observed births. 

# USA data

USA _male_ fertility data from Dudel, C. and S. Klüsener. Male fertility data for high-income 
countries [unpublished data]. Submitted to the HFC by C. Dudel # on 15.05.2019.
downloaded from Human Fertility Collection 2 Nov 2019
https://www.fertilitydata.org/data/RAW_DATA/m_USA_51.zip

USA _female_ fertility data from H**F**D

USA _male_ and _female_ population data from H**M**D

```{r}

# construct a data frame with annual USA data, from the listed sources

# rates for 1969-2015 x ages 15-59
asfr_male = read.table(file='m_ASFR_USA.txt', header=TRUE,
                       colClasses='numeric') %>%
            add_column(Sex='Male')

# rates for 1969-2015 x ages 12-55
asfr_female = read.table(file='USAasfrTR.txt', skip=2, header=TRUE) %>%
       mutate(Age = 11 + as.numeric(Age)) %>%
       filter(Year %in% unique(asfr_male$Year)) %>%
       group_by(Year,Age) %>%
       summarize(ASFR = mean(ASFR)) %>%
       add_column(Sex='Female')

ASFR = bind_rows(
         asfr_male,
         asfr_female
        ) 

pop = read.table('Population.txt',skip=2, header=TRUE) %>%
        mutate( Year = as.numeric(as.character(Year)),
                Age  = as.numeric(as.character(Age))) %>%
        filter(Age %in% unique(ASFR$Age),
               Year %in% unique(ASFR$Year)) %>%
        select(Year,Age,Female,Male) %>%
        gather(key=Sex, value=pop, -Year, -Age)

## add estimated births, fraction of births by age, etc.
 
big = full_join(pop, ASFR, by=c('Year','Age','Sex')) %>%
         mutate(births = pop * ASFR) 

big$births[is.na(big$births)] = 0

# summarize each year x sex 

mean_expos = function(A,P,M) {
  approx(x=A, y=P, xout=M )$y  
}

df = big %>%
       group_by(Year,Sex) %>%
       summarize( TFR = sum(ASFR, na.rm=TRUE),
                  B   = sum(births),
                  m   = weighted.mean( Age+.50   , births),
                  Nm  = mean_expos(Age+.50,pop,m),
                  approx_TFR = B/Nm)


ggplot(data=df) +
   aes(x=Year, y=TFR, color=Sex) +
   geom_line(lwd=1.5) +
   scale_color_manual(values=c('red','blue')) +
   labs(title='US Sex-Specific TFR 1969-2015') +
  theme_bw()

ggplot(data=df) +
   aes(x=Year, y=m, color=Sex) +
   geom_line(lwd=1.5) +
   scale_color_manual(values=c('red','blue')) +
   labs(title='US Mean Age of Parents at Birth 1969-2015') +
  theme_bw()

ggplot(data=df) +
   aes(x=Year, y=Nm, color=Sex) +
   geom_smooth(lwd=1.5,se=FALSE) +
   scale_color_manual(values=c('red','blue')) +
   labs(title='US Population at Mean Age at Birth 1969-2015 (smoothed)') +
  theme_bw()

```


# Example: US TFRs in `r detail_year`

Select `r detail_year` data and analyze in more detail. 

```{r}
tmp.big = filter(big, Year == detail_year)
tmp.df  = filter(df,  Year == detail_year)

TFR        = tmp.df$TFR
approx_TFR = tmp.df$approx_TFR
m          = tmp.df$m
Nm         = tmp.df$Nm

B = mean(tmp.df$B) #avg of (slightly difft) male and female total births

names(TFR) = names(approx_TFR) = names(m) = names(Nm) = c('Female','Male')

ratio        = TFR['Male']/TFR['Female']
approx_ratio = approx_TFR['Male']/approx_TFR['Female']

```

For the US in `r detail_year`, female TFR was **`r round(TFR['Female'],2)`** and male TFR was  **`r round(TFR['Male'],2)`**.

Now illustrate `r detail_year` fertility rates, births, and populations, by age and sex. 

```{r}

ggplot(data=tmp.big) +
  aes(x=Age, y=ASFR, color=Sex) +
  geom_line(lwd=1.5) +
  scale_color_manual(values=c('red','blue')) +
  labs(title=paste('US Age-Specific Fertility Rates',detail_year),
       subtitle=paste('Female TFR=',round(TFR['Female'],2),
                      '  Male TFR=',round(TFR['Male']  ,2))) +
  theme_bw()

  
ggplot(data=tmp.big) +
  aes(x=Age, y=births, color=Sex) +
  geom_line(lwd=1.5) +
  geom_vline(xintercept = m[c('Female','Male')], color=c('red','blue'),
             lwd=1.5, lty='dotted') +
  scale_color_manual(values=c('red','blue')) +
  labs(title=paste('US Age-Specific Births',detail_year),
       subtitle=paste('Mother\'s Mean Age =',round(m['Female'],2),
                      'Father\'s Mean Age =',round(m['Male'],2))) +
  theme_bw()

ggplot(data=tmp.big) +
  aes(x=Age+0.5, y=pop, color=Sex) +
  geom_line(lwd=1.5) +
  geom_vline(xintercept = m[c('Female','Male')], color=c('red','blue'),
             lwd=1.5,lty='dashed') +
  scale_color_manual(values=c('red','blue')) +
  labs(title=paste('US Age-Specific Populations',detail_year),
       x='Age') +
  theme_bw()

```

Now we will check the approximations based on

* mean parental ages for $`r detail_year`$ births:  $`r round(m['Female'],2)`$ for females and $`r round(m['Male'],2)`$ for males
* $`r detail_year`$ populations at mean ages:  $`r round(Nm['Female'])`$  for females and  $`r round(Nm['Male'])`$ for males 

Thus 

Sex   |  True `r detail_year` TFR  | m |   $N_m$ (000s) |  Approx TFR
----- | ------------------------- | --| -------------| -------------- 
Male    | `r round(TFR['Male'],2)` | `r round(m['Male'],2)` |  `r round(Nm['Male']/1000)` | `r round(approx_TFR['Male'],2)`
Female    | `r round(TFR['Female'],2)` | `r round(m['Female'],2)` |  `r round(Nm['Female']/1000)` | `r round(approx_TFR['Female'],2)`

and

True Male/Female TFR Ratio | Approximate Ratio
----------------------- | ------------------
`r round(ratio,2)` | `r round(approx_ratio,2)` 

------ 


## Appendix: Approximation

We can approximate the $TFR$s in (\ref{eq:mainTFR}) by considering a Taylor series approximation for $N^{-1}_x$. For the purposes of this derivation, temporarily ignore the discrete nature of ages in observed data, and treat $N_x$ as a continuous function of exact age $x$. 

Define the (period) mean age of parents in our group as $m$. The second-order Taylor approximation to $N^{-1}_x$ is 
$$
N^{-1}_x \approx N^{-1}_m + (x-m) \left[ -N^{-2}_m N'_m \right]
+ \tfrac{1}{2} (x-m)^2\left[ 2N^{-3}_m N'_m - N^{-2}_m N''_m\right]
$$
Thus the expectation in (\ref{eq:mainTFR}) is 
$$
E(N^{-1}_x) \approx N^{-1}_m  
+ \tfrac{\sigma^2_x}{2} \left[ 2N^{-3}_m N'_m - N^{-2}_m N''_m\right]
$$
where $\sigma^2_x$ is the observed variance of parental ages. Typical values for the variance term are in $[25,50]$, so when exposure $N$ is measured in thousands or millions, the second-order term is many orders of magnitude smaller than the first-order term. 

Using only the first term leads to a very simple approximation for $TFR$, based on (\ref{eq:mainTFR}) that is fairly robust:
$$
TFR \; \approx \; \frac{B}{N_m} = \frac{\text{total annual births}}{\text{mid-year population at mean parental age}}
$$
