---
title: "USA TFR by sex"
author: "Carl Schmertmann"
date: "6 Nov 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Notation

Suppose that there are $A$ reproductive ages $a=1\ldots A$ in which either men or women may have children. At some of these ages, notably 50+, male fertility may be possible even though female fertility is not. 

Suppose also that there are a total of $B$ births in a given year, and that $p_{xy}$ is the fraction of those births for which the mother is $x$ years old and the father is $y$ years old.  

Call the $\mathbf{P}$ the $A \times A$ matrix of $p_{xy}$ values, with mother's age corresponding to rows and father's age to columns. The marginal distribution of mothers' ages for this year's births is $\mathbf{P}\mathbf{1} \in \mathbb{R}^A$, and the marginal distribution of fathers' ages is $\mathbf{P}^\prime\mathbf{1} \in \mathbb{R}^A$, where $\mathbf{1}$ is an $A \times 1$ vector of ones. 

The vector of births by mother's age is thus
$$
\mathbf{b}_{female} = B \cdot \mathbf{P}\,\mathbf{1} \in \mathbb{R}^A
$$
and the vector of births by father's age is
$$
\mathbf{b}_{male} = B \cdot \mathbf{P}^\prime \,\mathbf{1} \in \mathbb{R}^A
$$
Define $\mathbf{W} = diag(W_1\ldots W_A)$ as a diagonal matrix with population counts of _women_, and $\mathbf{M} = diag(M_1\ldots M_A)$ as a diagonal matrix with population counts of _men_. 

Age-specific birth rates and $TFR$ for women are therefore
$$
\mathbf{f}_{female} = \mathbf{W}^{-1} \mathbf{b}_{female} = B \cdot  \mathbf{W}^{-1} \mathbf{P} \,\mathbf{1}
$$
and 
$$
TFR_{female} = \mathbf{1}^\prime \mathbf{f}_{female} =  B \cdot \mathbf{1}^\prime \mathbf{W}^{-1} \mathbf{P} \,\mathbf{1}
$$
Because $\mathbf{P}\mathbf{1} \in \mathbb{R}^A$ is the marginal distributions of mother's ages for this year's births, we can interpret this as 
$$
TFR_{female} = B\, \cdot \, E_{x} \left( \frac{1}{W_x}\right)
$$
where the expectation is weighted by mothers' ages of this year's births. 

An analogous derivation for men produces age-specific rates
$$
\mathbf{f}_{male} = \mathbf{M}^{-1} \mathbf{b}_{male} = B \cdot  \mathbf{M}^{-1} \mathbf{P}^\prime \,\mathbf{1}
$$
and 
$$
TFR_{male} = \mathbf{1}^\prime \mathbf{f}_{male} =  B \cdot \mathbf{1}^\prime \mathbf{M}^{-1} \mathbf{P}^\prime \,\mathbf{1}
$$
We can interpret this as 
$$
TFR_{male} = B\, \cdot \, E_{y} \left( \frac{1}{M_y}\right)
$$
where the expectation is weighted by fathers' ages of this year's births. 


The difference of the sex-specific $TFR$s is
$$
TFR_{male} - TFR_{female} = B \cdot \left[  E_{y} \left( \frac{1}{M_y}\right) - E_{x} \left( \frac{1}{W_x}\right)\right]
$$
In very broad terms, we can see that this difference is more likely to be positive (higher male TFR) if there are more women and fewer men in the reproductive-age population. More subtly, the age distributions of males and females also affect the difference in sex-specific $TFR$s. 

# Example: US TFRs in 1975

```{r}
# USA male fertility
# from Dudel, C. and S. Klüsener. Male fertility data for high-income 
# countries [unpublished data]. Submitted to the HFC by C. Dudel # on 15.05.2019.
# 
# downloaded from Human Fertility Collection 2 Nov 2019
# https://www.fertilitydata.org/data/RAW_DATA/m_USA_51.zip


# Male 1975 rates from HFC, as described above
asfr_male = read.table(file='m_ASFR_USA.txt', header=TRUE) %>%
              filter(Year=='1975') %>%
              pull(ASFR)

# male data pulled is for ages 15-59. Add zero rates for 12-14
asfr_male = c( rep(0,3), asfr_male)

# Female 1975 rates from HFD
asfr_female = read.table(file='USAasfrTR.txt', skip=2, header=TRUE) %>%
       mutate(Age = 11 + as.numeric(Age)) %>%
       group_by(Year,Age) %>%
       summarize(ASFR = mean(ASFR)) %>%
       filter(Year=='1975') %>%
       pull(ASFR)

# female data pulled is for ages 12-54. Add zero rates for 56-59
asfr_female = c( asfr_female, rep(0,4))

# 1975 US populations

pop = read.table('Population.txt',skip=2, header=TRUE) %>%
        filter(Year=='1975', Age %in% 12:59) %>%
        select(Age,Female,Male)

Bx = pop$Female * asfr_female
By = pop$Male   * asfr_male

px = prop.table(Bx)
py = prop.table(By)

female_factor = weighted.mean( 1/pop$Female, w=px)
male_factor   = weighted.mean( 1/pop$Male  , w=py)



```